Trifunctional Sphingosine

### 2025-01-16 LipidProbe/Phosphatidylethanolamine.qmd
Updates for Carsten

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Volcano plots depict the relative enrichment of each protein versus non-crosslinked control (x-axis) and the statistical significance of each protein following a Student's t-test (y-axis, -log10 transformed). The left panel depicts the results of a pulldown using trifunctionalized phosphatidic acid; the right panel depicts the results of the pulldown using trifunctionalized phosphatidylethanolamine. Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

@@

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Ranked-order plots depict the relative enrichment of each protein versus non-crosslinked control (y-axis) from lowest to highest. The left panel depicts the results of a pulldown using trifunctionalized phosphatidic acid; the right panel depicts the results of the pulldown using trifunctionalized phosphatidylethanolamine. Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

@@

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "MA plots depict the average abundance of each protein in the crosslinked and uncrosslinked conditions (x-axis) versus the log2-transformed fold-change between the crosslinked and uncrosslinked conditions (y-axis). Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

>>>>

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Data as reported in [Thomas et al., 2025](/IndividualStudies/AT_2025.qmd). Volcano plots depict the relative enrichment of each protein versus non-crosslinked control (x-axis) and the statistical significance of each protein following a Student's t-test (y-axis, -log10 transformed). The left panel depicts the results of a pulldown using trifunctionalized phosphatidic acid; the right panel depicts the results of the pulldown using trifunctionalized phosphatidylethanolamine. Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

@@

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Data as reported in [Thomas et al., 2025](/IndividualStudies/AT_2025.qmd). Ranked-order plots depict the relative enrichment of each protein versus non-crosslinked control (y-axis) from lowest to highest. The left panel depicts the results of a pulldown using trifunctionalized phosphatidic acid; the right panel depicts the results of the pulldown using trifunctionalized phosphatidylethanolamine. Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

@@

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Data as reported in [Thomas et al., 2025](/IndividualStudies/AT_2025.qmd). MA plots depict the average abundance of each protein in the crosslinked and uncrosslinked conditions (x-axis) versus the log2-transformed fold-change between the crosslinked and uncrosslinked conditions (y-axis). Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

### 2025-01-16 LipidProbe/Sphinganine.qmd
minor tweaks
added some PIP buttons, still need content
Updates for Carsten

---
title: "Trifunctional Phosphatidic Acid"
---

@@

>>>>

---
title: "Trifunctional sphinganine"
---

@@

### Cell background: Huh7

::: {.callout-note collapse="true"}

#### Study details

##### Authors
Scotland E. Farley, Frank Stein, Per Haberkant, Fikadu G. Tafesse, Carsten Schultz

##### Journal
ACS Chemical Biology

<https://doi.org/10.1021/acschembio.3c00554>{target="_blank"}

##### Abstract

Functions of the sphingolipids sphingosine and sphinganine in cells are not well established. While some signaling roles for sphingosine have been elucidated, the closely related sphinganine has been described only insofar as it does not elicit many of the same signaling responses. The underlying mechanisms behind the cell biological differences between these lipids are not well understood. Here, we prepared multifunctionalized derivatives of the two lipid species that only differ in a single double bond of the carbon backbone. Using these novel probes, we were able to define their spatiotemporal distribution within cells. Furthermore, we used these tools to systematically map the protein interactomes of both lipids. The lipid-protein conjugates, prepared through photo-crosslinking in live cells and extraction via click chemistry to azide beads, revealed significant differences in the captured proteins, highlighting their distinct roles in various cellular processes. This work elucidates mechanistic differences between these critical lipids and sets the foundation for further studies on the functions of sphingosine and sphinganine.

##### Lipid probes utilized

##### Cell line analyzed

[Huh7](https://huh7.com){target="_blank"}

##### Mass spectrometry quantification method

16-channel Tandem Mass Tagging (TMT16)

##### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
:::

::: {.panel-tabset}

```{r setup, echo=FALSE, warning=FALSE, include = FALSE}
## intitialize libraries
library(dplyr)
library(tidyr)
library(plotly)
library(ggplot2)
library(grid)
library(readr)
library(rmarkdown)
library(emojifont)
library(gt)
library(crosstalk)
library(DT)

## Set up the source for the standardized ggplot styles
source("ggplot_styles.R")

#  SF_2024_data <- readr::read_csv("LipidProbe/DataSets/SF_ACS-Chem-Bio_2024_download.csv") |>
#  	filter(LipidProbe == "Spa")

# 	write.csv(SF_2024_data, "LipidProbe/DataSets/Spa_Huh7_SF_2024.csv")

# Opening the datatable for this page
SF_2024_data <- readr::read_csv("DataSets/Spa_Huh7_SF_2024.csv") |>
	filter(LipidProbe == "Spa")

# Organizing the hit annotations as factors
SF_2024_data$hit_annotation <- factor(pull(SF_2024_data, hit_annotation), levels = c("enriched hit", "hit", "enriched candidate", "candidate", "no hit", "NA"), ordered = TRUE)

# Building an ID column to maintain the order on the x-axis
ID <- c()
for(i in 0:(nrow(SF_2024_data))){
	ID[i] = i
}

# Applies the ID column to the Spa results
SF_2024_ranked <- SF_2024_data |>
    arrange(logFC) |>
	filter(LipidProbe == "Spa") |>
	mutate(ID = ID)

```

##### Volcano plot

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Data as reported in [Farley et al., 2024](/IndividualStudies/SF_2024.qmd). Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

VolcanoPlotStandardized(SF_2024_data)
```
:::

##### Ranked order plot

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Data as reported in [Farley et al., 2024](/IndividualStudies/SF_2024.qmd). Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

RankedOrderPlotStandard(SF_2024_ranked)
```
:::

##### MA Plots

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Data as reported in [Farley et al., 2024](/IndividualStudies/SF_2024.qmd). Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

MAStandard(SF_2024_data)

```
:::
:::

Check the boxes below to filter by significance thresholds.

:::{.border}
```{r, echo=FALSE, warning=FALSE}

# Here is the code to make the filterable lists shown in the crosstalk table below

SF_2024_data <- SF_2024_data |>
  select(gene_name, LipidProbe, hit_annotation, logFC, pvalue, fdr) |>
  mutate(logFC = signif(logFC, digits=3),
          pvalue = signif(pvalue, digits=3),
          fdr = signif(fdr, digits = 3)) |>
  arrange(hit_annotation) |>
  arrange(LipidProbe)

## Creates a reactive dataframe
SF_2024_data_reactive <- crosstalk::SharedData$new(SF_2024_data)

htmltools::tags$div(
  class = "filter-container",
  filter_checkbox(id = "hit_annotation",
                  label = "Enrichment criteria:", 
                  sharedData = SF_2024_data_reactive,
                  group = ~hit_annotation)
)

## Displays the reactive dataframe
datatable(SF_2024_data_reactive, 
          options = list(pageLength = 10),
                         colnames = c("Gene Name", "Lipid Probe", "Hit Annotation", "LogFC", "p value", "FDR"))

```
:::

{{< downloadthis DataSets/Sph_Huh7_SF_2024.csv dname="Sph_Huh7_SF_ACS-Chem-Bio_download.csv">}}

### 2025-01-16 LipidProbe/Sphingosine.qmd
Updating individual study pages
checking render of PE page
Updates for Carsten

## Subcellular localization

@@

>>>>

![](/LipidProbe/Structures/SphingosineStructure.png)

@@

### Cell background: Huh7

::: {.callout-note collapse="true"}

#### Study details

##### Authors
Scotland E. Farley, Frank Stein, Per Haberkant, Fikadu G. Tafesse, Carsten Schultz

##### Journal
ACS Chemical Biology

<https://doi.org/10.1021/acschembio.3c00554>{target="_blank"}

##### Abstract

Functions of the sphingolipids sphingosine and sphinganine in cells are not well established. While some signaling roles for sphingosine have been elucidated, the closely related sphinganine has been described only insofar as it does not elicit many of the same signaling responses. The underlying mechanisms behind the cell biological differences between these lipids are not well understood. Here, we prepared multifunctionalized derivatives of the two lipid species that only differ in a single double bond of the carbon backbone. Using these novel probes, we were able to define their spatiotemporal distribution within cells. Furthermore, we used these tools to systematically map the protein interactomes of both lipids. The lipid-protein conjugates, prepared through photo-crosslinking in live cells and extraction via click chemistry to azide beads, revealed significant differences in the captured proteins, highlighting their distinct roles in various cellular processes. This work elucidates mechanistic differences between these critical lipids and sets the foundation for further studies on the functions of sphingosine and sphinganine.

##### Lipid probes utilized

##### Cell line analyzed

[Huh7](https://huh7.com){target="_blank"}

##### Mass spectrometry quantification method

16-channel Tandem Mass Tagging (TMT16)

##### Additional sample preparation <span class="help-symbol" title="e.g. sub-cellular fractionation, depletion, etc.">?</span>
:::

::: {.panel-tabset}

<!-- Data wrangling to make the plots -->

```{r setup, echo=FALSE, warning=FALSE, include = FALSE}
## intitialize libraries
library(dplyr)
library(tidyr)
library(plotly)
library(ggplot2)
library(grid)
library(readr)
library(rmarkdown)
library(emojifont)
library(gt)
library(crosstalk)
library(DT)

## Set up the source for the standardized ggplot styles
source("ggplot_styles.R")

#  SF_2024_data <- readr::read_csv("LipidProbe/DataSets/SF_ACS-Chem-Bio_2024_download.csv") |>
#  	filter(LipidProbe == "Sph")

# 	write.csv(SF_2024_data, "LipidProbe/DataSets/Sph_Huh7_SF_2024.csv")

# Opening the datatable for this page
SF_2024_data <- readr::read_csv("DataSets/Sph_Huh7_SF_2024.csv") |>
	filter(LipidProbe == "Sph")

# Organizing the hit annotations as factors
SF_2024_data$hit_annotation <- factor(pull(SF_2024_data, hit_annotation), levels = c("enriched hit", "hit", "enriched candidate", "candidate", "no hit", "NA"), ordered = TRUE)

# Building an ID column to maintain the order on the x-axis
ID <- c()
for(i in 0:(nrow(SF_2024_data))){
	ID[i] = i
}

# Applies the ID column to the Spa results
SF_2024_ranked <- SF_2024_data |>
    arrange(logFC) |>
	filter(LipidProbe == "Sph") |>
	mutate(ID = ID)

```

##### Volcano plot

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Data as reported in [Farley et al., 2024](/IndividualStudies/SF_2024.qmd). Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

VolcanoPlotStandardized(SF_2024_data)
```
:::

##### Ranked order plot

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Data as reported in [Farley et al., 2024](/IndividualStudies/SF_2024.qmd). Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

RankedOrderPlotStandard(SF_2024_ranked)
```
:::

##### MA Plots

:::{.border}
```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Data as reported in [Farley et al., 2024](/IndividualStudies/SF_2024.qmd). Black proteins are unenriched or depleted in the presence of probe, Purple enriched candidates are defined as proteins with a false discovery rate less than 0.2 and a fold change of at least 1.5-fold, and Orange enriched hits are defined as proteins with a false discovery rate less than 0.05 and a fold change of at least 2-fold in the +UV over the -UV)."

MAStandard(SF_2024_data)

```
:::
:::

Check the boxes below to filter by significance thresholds.

:::{.border}
```{r, echo=FALSE, warning=FALSE}

# Here is the code to make the filterable lists shown in the crosstalk table below

SF_2024_data <- SF_2024_data |>
  select(gene_name, LipidProbe, hit_annotation, logFC, pvalue, fdr) |>
  mutate(logFC = signif(logFC, digits=3),
          pvalue = signif(pvalue, digits=3),
          fdr = signif(fdr, digits = 3)) |>
  arrange(hit_annotation) |>
  arrange(LipidProbe)

## Creates a reactive dataframe
SF_2024_data_reactive <- crosstalk::SharedData$new(SF_2024_data)

htmltools::tags$div(
  class = "filter-container",
  filter_checkbox(id = "hit_annotation",
                  label = "Enrichment criteria:", 
                  sharedData = SF_2024_data_reactive,
                  group = ~hit_annotation)
)

## Displays the reactive dataframe
datatable(SF_2024_data_reactive, 
          options = list(pageLength = 10),
                         colnames = c("Gene Name", "Lipid Probe", "Hit Annotation", "LogFC", "p value", "FDR"))

```
:::

{{< downloadthis DataSets/Sph_Huh7_SF_2024.csv dname="Sph_Huh7_SF_ACS-Chem-Bio_download.csv">}}

### Cell background: HeLa

::: {.callout-note collapse="true"}

#### Study details

##### Authors

Doris Höglinger, André Nadler,  Per Haberkant, Joanna Kirkpatrick, Martina Schifferer, Frank Stein, Sebastian Hauke, Forbes D. Porter, Carsten Schultz

##### Journal

PNAS

<https://doi.org/10.1073/pnas.1611096114>{target="_blank"}

##### Abstract

Lipid-mediated signaling events regulate many cellular processes. Investigations of the complex underlying mechanisms are difficult because several different methods need to be used under varying conditions. Here we introduce multifunctional lipid derivatives to study lipid metabolism, lipid−protein interactions, and intracellular lipid localization with a single tool per target lipid. The probes are equipped with two photoreactive groups to allow photoliberation (uncaging) and photo–cross-linking in a sequential manner, as well as a click-handle for subsequent functionalization. We demonstrate the versatility of the design for the signaling lipids sphingosine and diacylglycerol; uncaging of the probe for these two species triggered calcium signaling and intracellular protein translocation events, respectively. We performed proteomic screens to map the lipid-interacting proteome for both lipids. Finally, we visualized a sphingosine transport deficiency in patient-derived Niemann−Pick disease type C fibroblasts by fluorescence as well as correlative light and electron microscopy, pointing toward the diagnostic potential of such tools. We envision that this type of probe will become important for analyzing and ultimately understanding lipid signaling events in a comprehensive manner.

##### Lipid probes utilized

Sphinganine

Diacylglycerol

8-3 Fatty acid

##### Cell line analyzed

[HeLa](https://www.atcc.org/products/ccl-2){target="_blank"}

##### Mass spectrometry quantification method

Peptide spectral counting

:::
